An Experimental Evaluation of Hybrid Vector Search (EA\&B)

### 2025-04-14 VLDB2026/main.tex
range filter pic

	\textit{\textbf{Multi-Attribute Building and Single-Attribute Query.}}
	In this scenario, the index is constructed using multiple attributes but applies filtering condition on only one attribute during the query process.

	In database systems (VBASE, PASE, Milvus), when the system builds both B+ tree and vector indexes, queries typically utilize only the B+ tree. Faiss offers an ID filtering mechanism, it operates solely during query execution and does not influence index building. ACORN-1 and ACORN-\(\gamma\) are specifically designed for single-attribute indexing. Hence, they are excluded from this comparison.

@@

	\textit{Effect of Multi-Attribute Building on Algorithm Performance (M vs. S).}
	As shown in Figure \ref{fig:exp_2_1}, Puck and CAPS demonstrate nearly no performance degradation. In contrast, the performance of FilteredVamana,StitchedVamana, NHQ, and UNG drop significantly, with the following reasons: 1) FilteredVamana increases the complexity introduced by retaining attribute during index construction. 2) The complexity of StitchedVamana increases because data points may duplicate across multiple subgraphs. 3) The fusion distance of NHQ exhibits sensitivity to the number of attributes, which degrades its effectiveness. 4) The reason why UNG has poor performance is that each base vector has 8 attributes, but only one query attribute, which results in a large number of entry points that need to be traversed. It is difficult for a graph with multiple entries to have good connectivity, degenerating into a situation closer to inverted retrieval, and the effect may be greatly impaired.

	\textit{Performance of Algorithms under Multi-Attribute Building (M vs. M).}  
	As shown in Figur \ref{fig:exp_2_1}, CAPS has the best QPS performance, while Puck has better recall accuracy. CAPS first partitions by clustering, and then builds AFT under each partition. For CAPS, whether it is built with multiple attributes or a single attribute, the number of points required to be retrieved will not change after AFT filtering during search, so the performance is not greatly affected. Puck filters out irrelevant results during retrieval rather than searching for more candidates. FilteredVamana performs the worst because its multi-attribute indexing forces neighbor selection to cover multiple attributes, hurting single-attribute query efficiency. This leads to excessive irrelevant node traversal, making it slower than other methods.

@@

	\begin{table}[t]
		\centering
		\setlength{\abovecaptionskip}{0.05cm}
		\caption{Time and space overhead of range filtering index construction}

		\label{tab:Range Filtered}
		\resizebox{\columnwidth}{!}{
			\begin{tabular}{l l c c c c c }
				\toprule
				\textbf{Dataset} & \textbf{Metric} & \textbf{DSG} & \textbf{iRange} & \textbf{SeRF} & \textbf{UNIFY} & \textbf{WinFilter} \\
				\midrule
				\multirow{3}{*}{DEEP} 
				& Index time(s)  & 4156 & 1128 & \textbf{189.89} & 7510 & 32843 \\
				& Index size(GB)  &  4.6 & 0.92 & \textbf{0.80} & 1.43 & 2.41  \\
				& Peak memory(GB) &  7.78 & 3.54 & \textbf{1.38} & 2.61 & 7.08   \\
				% \cmidrule{1-7}
				\midrule
				\multirow{3}{*}{YT-Audio} 
				& Index time(s)   & 15314 & 1039 & \textbf{187.20} & 5905 & 30904    \\
				& Index size(GB)    & 4.3 & \textbf{0.7} & 1.04 & 1.55 & 1.32  \\
				& Peak memory(GB)   & 7.5 & 3.16 & \textbf{1.57} & 2.97 & 6.95  \\
				\midrule
				\multirow{3}{*}{WIT} 
				& Index time(s)  & 29492 & 7068 & \textbf{828.12} & 28759 & 94909\\
				& Index size(GB) &  4.2 & \textbf{1.5 }& 15.34 & 8.71 & 1.59  \\
				& Peak memory(GB)  & 21.7 & \textbf{12.32} & 15.87 & 24.38 & 15.29\\
				% \cmidrule{1-7}
				\bottomrule
			\end{tabular}
		}
	\end{table}

@@

	\subsection{Range Filtering}
	Although the libraries and databases mentioned above can also implement range filtering, the results are poor, so we only compare 5 range-filtering algorithms.
	\subsubsection{Time and Space overhead of index construction.}

@@

	\textit{\textbf{Index construction time.}}
	As shown in Table \ref{tab:Range Filtered}, SeRF achieves the fastest index construction time among the 5 algorithms. It incrementally builds the index in attribute order. It requires only a single pass and adds edges within local graphs. The structure is simple, with no redundant operations.

@@

%	 修改此处
	ACORN performs similarly on both large and small range queries, but its overall performance is worse than the other methods.

@@

	To examine the robustness of RF-ANNS algorithms, we analyze their performance across different datasets. As shown in Figure~\ref{fig:exp_8_3}, the performance differences. All methods perform worst on WIT, which has the highest LID. On Deep and YT-Audio, which have similar LID, most algorithms achieve comparable performance. 
	This suggests that the effectiveness of RF-ANN search is mainly influenced by LID of the dataset. A higher LID inherently increases the difficulty of hybrid query.

	However, the DSG algorithm shows a noticeable performance gap between the Deep and YT-Audio datasets, despite their comparable LID. This indicates that DSG is less robust than other methods when facing dataset-dependent variations, possibly due to its sensitivity to data distribution beyond LID alone.

>>>>

	\textit{\textbf{Multi-Attribute Building and Single-Attribute Query.}}
	In this scenario, the index is constructed using 8 attributes but applies filtering condition on only 1 attribute during the query process.

	 Faiss offers an ID filtering mechanism, it operates solely during query execution and does not influence index building. ACORN-1 and ACORN-\(\gamma\) are specifically designed for single-attribute indexing. In database systems (VBASE, PASE, Milvus), when the system builds both B+ tree and vector indexes, queries typically utilize only the B+ tree. Hence, they are excluded from this comparison.

@@

	\textit{Effect of Multi-Attribute Building on Algorithm Performance (M vs. S).}
	As shown in Figure \ref{fig:exp_2_1}, Puck demonstrates nearly no performance degradation. In contrast, CAPS shows a slight performance decline due to the increased query overhead caused by excessive attribute-based grouping. Compared with CAPS, the performance of StitchedVamana, FilteredVamana, NHQ, and UNG drops significantly, with the following reasons: 1) FilteredVamana increases the complexity introduced by retaining attribute during index construction. 2) The complexity of StitchedVamana increases because data points may duplicate across multiple subgraphs. 3) The fusion distance of NHQ exhibits sensitivity to the number of attributes, which degrades its effectiveness. 4) UNG has poor performance because large number of entry points that need to be traversed. It is difficult for a graph with multiple entries to have good connectivity, degenerating into a situation closer to inverted retrieval, and the effect may be greatly impaired.

	\textit{Performance of Algorithms under Multi-Attribute Building (M vs. M).}  
	As shown in Figure \ref{fig:exp_2_1}, IVF-based methods (Puck and CAPS) significantly outperform graph-based methods. The reason for Puck's superior performance is that it filters out irrelevant results during the retrieval process rather than searching for more candidates. The performance of CPAS is as good as that of Puck. The difference between them is that CPAS further groups the results by attributes within clusters. In contrast, FilteredVamana performs the worst because its multi-attribute indexing forces neighbor selection to cover multiple attributes, thereby reducing the efficiency of single-attribute queries. This results in excessive traversal of irrelevant nodes, making it slower than other methods.

@@

	% \begin{table}[t]
	% 	\centering
	% 	\setlength{\abovecaptionskip}{0.05cm}
	% 	\caption{Time and space overhead of range filtering index construction}

	% 	\label{tab:Range Filtered}
	% 	\resizebox{\columnwidth}{!}{
	% 		\begin{tabular}{l l c c c c c }
	% 			\toprule
	% 			\textbf{Dataset} & \textbf{Metric} & \textbf{DSG} & \textbf{iRange} & \textbf{SeRF} & \textbf{UNIFY} & \textbf{WinFilter} \\
	% 			\midrule
	% 			\multirow{3}{*}{DEEP} 
	% 			& Index time(s)  & 4156 & 1128 & \textbf{189.89} & 7510 & 32843 \\
	% 			& Index size(GB)  &  4.6 & 0.92 & \textbf{0.80} & 1.43 & 2.41  \\
	% 			& Peak memory(GB) &  7.78 & 3.54 & \textbf{1.38} & 2.61 & 7.08   \\
	% 			% \cmidrule{1-7}
	% 			\midrule
	% 			\multirow{3}{*}{YT-Audio} 
	% 			& Index time(s)   & 15314 & 1039 & \textbf{187.20} & 5905 & 30904    \\
	% 			& Index size(GB)    & 4.3 & \textbf{0.7} & 1.04 & 1.55 & 1.32  \\
	% 			& Peak memory(GB)   & 7.5 & 3.16 & \textbf{1.57} & 2.97 & 6.95  \\
	% 			\midrule
	% 			\multirow{3}{*}{WIT} 
	% 			& Index time(s)  & 29492 & 7068 & \textbf{828.12} & 28759 & 94909\\
	% 			& Index size(GB) &  4.2 & \textbf{1.5 }& 15.34 & 8.71 & 1.59  \\
	% 			& Peak memory(GB)  & 21.7 & \textbf{12.32} & 15.87 & 24.38 & 15.29\\
	% 			% \cmidrule{1-7}
	% 			\bottomrule
	% 		\end{tabular}
	% 	}
	% \end{table}

@@

	\subsection{Range Filtering}
	Although the libraries and databases mentioned above can also implement range filtering, the results are poor, so we only compare 5 range-filtering algorithms.
	\subsubsection{Time and Space overhead of index construction.}
	%4.2.1

	\begin{figure}[t]
		\centering
		\setlength{\abovecaptionskip}{0.3cm}
		\setlength{\belowcaptionskip}{0cm}

		% 放大的图例（占满单栏）
		\includegraphics[width=0.98\columnwidth]{figures/indexData/rangeFilter_legend_only.pdf}

		\vspace{0.3cm}

		% 三图横排
		\begin{subfigure}[t]{0.31\columnwidth}
			\centering
			\includegraphics[width=\linewidth]{figures/indexData/rangeFilter_build_time_comparison_query.pdf}
			\caption{Index construction time}
			\label{fig:rangeFilter_build_time}
		\end{subfigure}
		\hfill
		\begin{subfigure}[t]{0.31\columnwidth}
			\centering
			\includegraphics[width=\linewidth]{figures/indexData/rangeFilter_index_size_mb_comparison_query.pdf}
			\caption{Index size}
			\label{fig:rangeFilter_ndex_size_mb}
		\end{subfigure}
		\hfill
		\begin{subfigure}[t]{0.31\columnwidth}
			\centering
			\includegraphics[width=\linewidth]{figures/indexData/rangeFilter_memory_mb_comparison_query.pdf}
			\caption{Peak memory usage}
			\label{fig:rangeFilter_memory_mb}
		\end{subfigure}

		\caption{Time and space overhead of range filtering index construction}
		\label{fig:rangeFilter_build_index_comparison}
	\end{figure}

@@

	\textit{\textbf{Index construction time.}}
, SeRF achieves the fastest index construction time among the 5 algorithms. It incrementally builds the index in attribute order. It requires only a single pass and adds edges within local graphs. The structure is simple, with no redundant operations.

@@

%	 修改此处
	ACORN performs similarly on both large and small range queries, but its overall performance is worse than the other methods. Unlike other range-filtering algorithms, ACORN is not designed with specialized data structures  to achieve efficient range filtering. In fact, the graph structure used by ACORN for attribute filtering is the same as that used for range filtering.

@@

To evaluate the robustness of the RF-ANNS algorithm in different scenarios, we added an Out-of-Distribution (OOD) dataset to the three standard datasets and tested performance under large-range queries. As shown in Figure~\ref{fig:exp_8_3}, most methods performed worst on the Text2Image dataset. This is because OOD queries make the $k$-nearest neighbor distribution more sparse. As a result, the search needs to visit more nodes, which leads to an exponential growth in the search space and a sharp drop in efficiency.

WinFilter showed the worst performance on the WIT dataset. This suggests that its performance is sensitive to the data dimensionality. ACORN had large performance differences across datasets, indicating weak robustness.

Among vector databases, Milvus performed well. PASE was not included in this round of tests because it does not support datasets with dimensions over 512 and performed poorly in earlier experiments.

It is worth noting that most methods achieved similar performance on the Deep and YT-Audio datasets, which have similar LID values. This shows that RF-ANNS performance is largely affected by both the LID and the dimensionality of the dataset. A high LID and high dimensionality make hybrid queries more difficult.

However, even though Deep and YT-Audio have similar LID values, the DSG algorithm showed large performance differences between them. This means that DSG is more sensitive to changes in data distribution. Its robustness is lower than that of other methods, possibly because it depends not only on LID but also on other complex data characteristics.
%	To examine the robustness of RF-ANNS algorithms, we analyze their performance across different datasets. As shown in Figure~\ref{fig:exp_8_3}, the performance differences. All methods perform worst on WIT, which has the highest LID. On Deep and YT-Audio, which have similar LID, most algorithms achieve comparable performance. 
%	This suggests that the effectiveness of RF-ANN search is mainly influenced by LID of the dataset. A higher LID inherently increases the difficulty of hybrid query.
%	
%	However, the DSG algorithm shows a noticeable performance gap between the Deep and YT-Audio datasets, despite their comparable LID. This indicates that DSG is less robust than other methods when facing dataset-dependent variations, possibly due to its sensitivity to data distribution beyond LID alone.
